---
title: "Assignment 3"
author: "Madimetja Maredi"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---


# Setup chunk

This chunk loads necessary R packages:

tidyverse: Collection of packages for data manipulation and visualisation

scales: Provides tools for formatting axis labels and values

corrplot: For correlation matrix visualisation

rstan: R interface to Stan for Bayesian modelling

bayesplot: For Bayesian model diagnostic plots
```{r setup, warning=FALSE, message=FALSE}
library("tidyverse")
library("scales")
library("corrplot")
library("rstan")
library("bayesplot")
```

# Data import

This chunk reads the AI companies dataset from a CSV file.
```{r data_import, message=FALSE}
data <- read_csv("Ai_companies.csv")
```

# Data Cleaning

This chunk performs comprehensive data cleaning:

It defines two helper functions:

clean_numeric() to convert text values like "10 million" to actual numbers
clean_rating() to standardize Glassdoor ratings

The main data transformation:

Selects only the columns needed for analysis

Renames columns for easier reference

Applies the cleaning functions to standardise data formats

Creates log transformations of Annual Revenue

Calculates company age by subtracting founding year from 2024

Creates log transformation of company age

Filters out entries with missing Glassdoor scores
```{r data_cleaning, message=FALSE, warning=FALSE}
clean_numeric <- function(x) {
  multiplier <- ifelse(grepl("million", x, ignore.case = TRUE), 1e6,
                       ifelse(grepl("billion", x, ignore.case = TRUE), 1e9, 1))
  num <- as.numeric(gsub("[^-0-9.]", "", x))
  num * multiplier
}

clean_rating <- function(x) {
  s <- ifelse(x == "5-Apr", "4", x)
  as.numeric(sub("/5", "", s))
}

Ass_data <- data %>%
  select(Founded, `Annual Revenue`, `Glassdoor Score`) %>%
  rename(Glass_Score = `Glassdoor Score`, Annual_Revenue = `Annual Revenue`) %>%
  mutate(
    Glass_Score = clean_rating(Glass_Score),
    Annual_Revenue = clean_numeric(Annual_Revenue),
    Annual_Revenue = parse_number(dollar(Annual_Revenue)),
    log_AR = log(Annual_Revenue),
    Age = 2024 - Founded,
    log_Age = log(Age)
  ) %>%
  filter(!is.na(Glass_Score))
```

# Data Quality Check Chunk

This chunk performs quality checks on the cleaned data:

Counts missing values across all columns

Creates boxplots to identify potential outliers in Annual Revenue and Glassdoor Score

Produces summary statistics for all variables

Generates histograms to visualize the distributions of Glassdoor Scores and log-transformed Annual Revenue
```{r}
missing_values <- Ass_data %>% 
  summarize(across(everything(), ~sum(is.na(.))))
print("Missing values check:")
print(missing_values)


ggplot(Ass_data, aes(y = Annual_Revenue)) + 
  geom_boxplot() + 
  labs(title = "Checking for outliers in Annual Revenue")

ggplot(Ass_data, aes(y = Glass_Score)) + 
  geom_boxplot() + 
  labs(title = "Checking for outliers in Glassdoor Score")

summary_stats <- Ass_data %>%
  summary()
print("Summary statistics after cleaning:")
print(summary_stats)


ggplot(Ass_data, aes(x = Glass_Score)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Glassdoor Scores after cleaning",
       x = "Glassdoor Score",
       y = "Count") +
  theme_minimal()

ggplot(Ass_data, aes(x = log_AR)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "black") +
  labs(title = "Distribution of log-transformed Annual Revenue after cleaning",
       x = "Log Annual Revenue",
       y = "Count") +
  theme_minimal()
```

The distribution of Glassdoor Scores clusters around higher ratings (3.5–4.5), indicating that most AI companies in our dataset tend to have good employee satisfaction. Meanwhile, the log-transformed Annual Revenue follows a more normal distribution than the raw values, validating our decision to apply the transformation for regression analysis.

# Data slicing chunk

This chunk partitions the data:

Extracts the first row as MyPos for later prediction purposes

Creates Ass_datanew dataset with all rows except the first one for model training

Prints the extracted data point that will be used for prediction
```{r data_slicing, message=FALSE, warning=FALSE}
MyPos <- Ass_data %>% slice(1)
print("Data point for prediction:")
print(MyPos)

Ass_datanew <- Ass_data %>% slice(-1)
```

# Exploratory Data Analysis Chunk

This chunk explores relationships between variables:

Creates scatter plots with fitted regression lines for:

Glassdoor Score vs. log Annual Revenue
log Age vs. log Annual Revenue


Calculates correlation coefficients between key variables

Visualises the correlation matrix using circles

Runs linear regression models to test statistical significance
```{r message=FALSE}
ggplot(Ass_datanew, aes(x = Glass_Score, y = log_AR)) +
  geom_point(alpha = 0.7, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Relationship between Glassdoor Score and log Annual Revenue",
       x = "Glassdoor Score", 
       y = "Log Annual Revenue") +
  theme_light()

ggplot(Ass_datanew, aes(x = log_Age, y = log_AR)) +
  geom_point(alpha = 0.7, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Relationship between log Age and log Annual Revenue",
       x = "Log Age", 
       y = "Log Annual Revenue") +
  theme_light()

correlation_matrix <- cor(Ass_datanew[, c("log_AR", "log_Age", "Glass_Score")])
print("Correlation matrix:")
print(correlation_matrix)

corrplot(correlation_matrix, method = "circle", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Key Variables")

lm_test1 <- lm(log_AR ~ Glass_Score, data = Ass_datanew)
lm_test2 <- lm(log_AR ~ log_Age, data = Ass_datanew)
print("Statistical significance of Glass_Score:")
print(summary(lm_test1)$coefficients)
print("Statistical significance of log_Age:")
print(summary(lm_test2)$coefficients)
```

Our analysis shows a moderate positive correlation (r = 0.51) between log Annual Revenue and log Age, suggesting that older AI companies tend to generate higher revenue. In contrast, the Glassdoor Score has a weaker correlation (r = 0.23) with log Annual Revenue, implying that employee satisfaction ratings may have a less significant impact on financial performance in this industry.


# Bayesian Model Fitting Chunk

This chunk defines and fits a robust Bayesian model:

Specifies a Stan model with Student-t likelihood (robust to outliers)

Defines model components:

Data section: Specifies input data structure

Parameters section: Defines model parameters (intercept, coefficients, error term, degrees of freedom)

Model section: Specifies priors and likelihood

Generated quantities: Creates posterior predictive samples


Prepares data for Stan

Fits the model with 4 chains and 2000 iterations
```{r robust_model, message=FALSE, warning=FALSE}

robust_model_code <- "
data {
  int N;               // Number of observations
  vector[N] y;         // Response variable (log-transformed Annual Revenue)
  int K;               // Number of predictors
  matrix[N, K] X;      // Predictor matrix (Glassdoor Score and log Age)
}
parameters {
  real alpha;          // Intercept
  vector[K] beta;      // Coefficients for predictors
  real<lower=0> sigma; // Scale (error) term
  real<lower=1> nu;    // Degrees of freedom (for Student-t likelihood)
}
model {
  // Priors for intercept and coefficients (using Student-t for robustness)
  alpha ~ student_t(7, 0, 5);
  beta ~ student_t(7, 0, 5);
  sigma ~ normal(0, 5);
  nu ~ gamma(2, 0.1);  // Prior for degrees of freedom
  
  // Likelihood with Student-t error distribution
  y ~ student_t(nu, alpha + X * beta, sigma);
}
generated quantities {
  // Generate posterior predictive samples for model checking
  vector[N] y_pred;
  for (n in 1:N)
    y_pred[n] = student_t_rng(nu, alpha + X[n] * beta, sigma);
}
"

robust_data <- list(
  N = nrow(Ass_datanew),
  y = Ass_datanew$log_AR,
  K = 2,
  X = as.matrix(Ass_datanew[, c("Glass_Score", "log_Age")])
)


robust_fit <- stan(
  model_code = robust_model_code,
  data = robust_data,
  iter = 2000,
  chains = 4,
  seed = 2014095653,
  refresh = 0  
)
```

# Model summary and interpretation chunk

This chunk provides comprehensive model interpretation:

Summarises model parameters

Labels coefficients for clarity

Extracts posterior samples for further analysis

Provides plain-language interpretation of coefficients

Visualises posterior distributions with histograms

Calculates and reports 95% credible intervals

Performs posterior predictive checks to validate model fit
```{r model_summary}

print("Robust Bayesian Model Summary:")
print(summary(robust_fit, pars = c("alpha", "beta", "sigma", "nu"))$summary)


robust_coefs <- summary(robust_fit, pars = c("alpha", "beta"))$summary
rownames(robust_coefs)[2:3] <- c("Glass_Score (beta[1])", "log_Age (beta[2])")
print("Coefficients with labeled parameters:")
print(robust_coefs)


posterior_samples <- extract(robust_fit)


glass_score_effect <- mean(posterior_samples$beta[,1])
log_age_effect <- mean(posterior_samples$beta[,2])
intercept <- mean(posterior_samples$alpha)
nu_value <- mean(posterior_samples$nu)

print("Interpretation of coefficients:")
print(paste("1. Glass_Score: On average, a 1-point increase in Glassdoor score is associated with a", 
    round(glass_score_effect, 3), "increase in log Annual Revenue, holding log Age constant."))
print(paste("2. log_Age: On average, a 1-unit increase in log Age is associated with a", 
    round(log_age_effect, 3), "increase in log Annual Revenue, holding Glassdoor score constant."))
print(paste("3. The degrees of freedom parameter (nu) is estimated at", round(nu_value, 2), 
           "suggesting a moderately heavy-tailed error distribution."))


par(mfrow=c(2,2))
hist(posterior_samples$alpha, main="Posterior for Intercept", xlab="alpha")
hist(posterior_samples$beta[,1], main="Posterior for Glass_Score coefficient", xlab="beta[1]")
hist(posterior_samples$beta[,2], main="Posterior for log_Age coefficient", xlab="beta[2]")
hist(posterior_samples$nu, main="Posterior for Degrees of Freedom", xlab="nu")
par(mfrow=c(1,1))


alpha_ci <- quantile(posterior_samples$alpha, probs = c(0.025, 0.975))
glass_score_ci <- quantile(posterior_samples$beta[,1], probs = c(0.025, 0.975))
log_age_ci <- quantile(posterior_samples$beta[,2], probs = c(0.025, 0.975))

print("95% Credible intervals:")
print(paste("Intercept:", round(alpha_ci[1], 3), "to", round(alpha_ci[2], 3)))
print(paste("Glass_Score effect:", round(glass_score_ci[1], 3), "to", round(glass_score_ci[2], 3)))
print(paste("log_Age effect:", round(log_age_ci[1], 3), "to", round(log_age_ci[2], 3)))


y_pred <- extract(robust_fit)$y_pred
ppc_dens_overlay(Ass_datanew$log_AR, y_pred[1:50,]) +
  ggtitle("Posterior predictive check")


```

# Prediction Chunk

This chunk performs predictive analysis:

Creates a modified scenario (decreasing Glassdoor score by 0.5 and increasing age by 1 year)

Defines a custom prediction function using posterior samples

Generates predictions for both original and modified scenarios

Visualises predictive distributions with density plots

Calculates and reports predictions in both log and dollar scales

Computes percentage change in predicted revenue

Calculates probability that revenue is higher in the modified scenario

Creates and formats a comprehensive prediction summary table with credible intervals
```{r predictions}

newdata <- MyPos %>%
  mutate(
    Glass_Score = Glass_Score - 0.5, 
    Age = Age + 1,                    
    log_Age = log(Age)                
  ) 


predict_function <- function(data_row, samples) {
  n_draws <- length(samples$alpha)
  preds <- samples$alpha +
           samples$beta[,1] * data_row$Glass_Score +
           samples$beta[,2] * data_row$log_Age
  return(preds)
}


pred_original <- predict_function(MyPos, posterior_samples)
pred_modified <- predict_function(newdata, posterior_samples)


pred_data <- data.frame(
  Modified = pred_modified,
  Original = pred_original
) %>%
  pivot_longer(cols = everything(), names_to = "Scenario", values_to = "Predicted_log_AR")

ggplot(pred_data, aes(x = Predicted_log_AR, fill = Scenario)) +
  geom_density(alpha = 0.5) +
  labs(title = "Posterior Predictive Distributions",
       subtitle = "Comparing original data point vs. modified scenario",
       x = "Predicted log Annual Revenue",
       y = "Density") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")


median_pred_orig <- median(pred_original)
median_pred_mod <- median(pred_modified)

cat("\nMedian predictions (log scale):\n")
cat("Original scenario:", median_pred_orig, "\n")
cat("Modified scenario:", median_pred_mod, "\n")

cat("\nMedian predictions (dollar amounts):\n")
cat("Original scenario:", dollar(exp(median_pred_orig)), "\n")
cat("Modified scenario:", dollar(exp(median_pred_mod)), "\n")

# Calculate and report percent change
pct_change <- (exp(median_pred_mod) - exp(median_pred_orig)) / exp(median_pred_orig) * 100

cat("\nPercent change in predicted revenue:", round(pct_change, 2), "%\n")


prob_higher <- mean(pred_modified > pred_original)
cat("\nProbability that revenue is higher in the modified scenario:", round(prob_higher * 100, 1), "%\n")


pred_summary <- data.frame(
  Scenario = c("Original", "Modified"),
  Median_Log_Revenue = c(median_pred_orig, median_pred_mod),
  Mean_Log_Revenue = c(mean(pred_original), mean(pred_modified)),
  Lower_CI_Log = c(quantile(pred_original, 0.025), quantile(pred_modified, 0.025)),
  Upper_CI_Log = c(quantile(pred_original, 0.975), quantile(pred_modified, 0.975)),
  Median_Dollar = c(exp(median_pred_orig), exp(median_pred_mod)),
  Lower_CI_Dollar = c(exp(quantile(pred_original, 0.025)), exp(quantile(pred_modified, 0.025))),
  Upper_CI_Dollar = c(exp(quantile(pred_original, 0.975)), exp(quantile(pred_modified, 0.975)))
)


pred_summary$Median_Dollar_Fmt <- dollar(pred_summary$Median_Dollar)
pred_summary$Lower_CI_Dollar_Fmt <- dollar(pred_summary$Lower_CI_Dollar)
pred_summary$Upper_CI_Dollar_Fmt <- dollar(pred_summary$Upper_CI_Dollar)

print("Comprehensive prediction summary:")
print(pred_summary[, c("Scenario", "Median_Log_Revenue", "Lower_CI_Log", "Upper_CI_Log", 
                       "Median_Dollar_Fmt", "Lower_CI_Dollar_Fmt", "Upper_CI_Dollar_Fmt")])
```

# My final analysis:

analysis employed a robust Bayesian approach with Student-t likelihood to model the 
relationship between company characteristics and annual revenue in AI companies. 
This model was specifically chosen for its ability to handle potential outliers and 
non-normal error distributions in the data.

Company Age Effect: The log-transformed company age is a strong predictor of revenue with a coefficient of  2.315. 
This suggests that older AI companies tend to generate significantly higher revenue, 
confirming the importance of market experience and business maturity.

Glassdoor Rating Effect: Employee satisfaction, as measured by Glassdoor scores, has a positive but more modest 
effect on revenue with a coefficient of  0.306. While the relationship is positive, 
it's substantially weaker than the effect of company age.


When we simulated decreasing the Glassdoor score by 0.5 while increasing the company age by 1 year, our model 
predicted a  net decrease  in revenue of  0.11 %. The probability that this change would lead to higher revenue is
49 %.

This prediction highlights the relative importance of these factors: 
the negative impact of a reduced Glassdoor score was  compounded  by the positive effect 
of increased company age.

This analysis provides strong evidence that company maturity is substantially more 
important for revenue growth than employee satisfaction ratings in AI companies. While Glassdoor 
scores do have a positive relationship with revenue, the effect is much smaller compared to the company's age.

The robust modelling approach with Student-t likelihood enhances our confidence in these findings by 
appropriately handling potential outliers and non-normality in the data. The heavy-tailed error distribution 
allows the model to accommodate unexpected variations that might be present in financial data.

For strategic decision-making, these results suggest that investors and stakeholders in 
AI companies should consider company maturity as a primary indicator of revenue potential, 
while still recognizing the positive but smaller contribution of employee satisfaction to financial performance.





# References

1. Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., & Rubin, D. B. (2013). Bayesian Data Analysis (3rd ed.). Chapman and Hall/CRC.

2. Johnson, A. A., Ott, M. Q., & Dogucu, M. (2022). Bayes Rules! An Introduction to Applied Bayesian Modeling. Chapman and Hall/CRC.

3. Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., ... & Riddell, A. (2017). Stan: A probabilistic programming language. Journal of Statistical Software, 76(1).

4. https://stackoverflow.com/questions/75759658/what-is-the-most-efficient-way-to-clean-currency-data-in-r